---
title: "Are we doing better at reducing the total loss in large earthquakes?"
author: "Chen Chen"
date: January 10, 2018
output: slidy_presentation
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Questions
<style type="text/css">
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}
</style>
1) Overall, are we doing better at reducing damages and fatality over the past 50 years? 

2) Are the total deaths/damages correlated with the magnitude/depth/location of the earthquakes?

3) Is there a correlation between a country's GDP/population and the total deaths/damages?

## Dataset

- [Significant earthquake database](https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1) from NOAA: damaging earthquakes from 2150 B.C. to present. 
- Earthquakes between 1967 and 2017 were downloaded, containing information on the earthquake data, location, depth, magnitude, total deaths, total damages, etc. 
- World Bank Database on Country's  [GDP](http://databank.worldbank.org/data/reports.aspx?source=2&series=NY.GDP.MKTP.CD&country=#) and [population](http://databank.worldbank.org/data/reports.aspx?source=2&series=SP.POP.TOTL&country=)

## Visualization of the Data 

```{r, message=FALSE, warning = FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggmap)
library(gridExtra)
library(grid)
library(randomForest)
library(cowplot)

# Read in and tidy the earthquake data 
eq <- read.table(file = 'results.tsv', sep = '\t', quote="\"",
                      header = TRUE, stringsAsFactors = F)

# Convert Date from character to date in R
eq_date <- eq %>% unite(DATE,YEAR:DAY,sep="-",remove=T) 
eq_date$DATE <- as.Date(eq_date$DATE)

# Reorder the columns and arrange the data frame by country 
country_eq <- eq_date %>% 
  select(COUNTRY:LONGITUDE,FOCAL_DEPTH,REGION_CODE,DATE:SECOND,everything()) %>% 
  arrange(COUNTRY)

# Next let's import and tidy the world gdp data
gdp <- read.csv(file='world_gdp_Data.csv',sep=",",
                col.names=c("Series.Name","Series.Code",
                "Country.Name","Country.Code",1960:2016),
                stringsAsFactors = F,check.names = F)

# Last 5 rows of gdp data table are explanation, and are excluded in the analysis
world_gdp <- head(gdp,-5)
gdp_tidy <- world_gdp %>% gather("Year","Value",5:61)
gdp_order <- gdp_tidy %>% select(Country.Name,Country.Code,everything(),-Series.Code) %>% 
  arrange(Country.Name)
gdp_spread <- gdp_order %>% spread(key=Series.Name,value = Value)
  
# Last let's tidy the population data 
popul <- read.csv(file="world_population_Data.csv",sep=",",
                       col.names=c("Series.Name","Series.Code",
                                   "Country.Name","Country.Code",1960:2016),
                  stringsAsFactors = F,check.names = F)

world_popu <- head(popul,-5)
popu_tidy <- world_popu %>% gather("Year","Population",5:61)
popu_order <- popu_tidy %>% arrange(Country.Name) %>% 
  select(Country.Name,Country.Code,Year,everything(),-Series.Code,-Series.Name)

# Add population in the gdp data table as a column.
gdp_popu <- left_join(gdp_spread,popu_order)

# Now the Fun Part: Perform data analysis 
gdp_popu[gdp_popu == ".."] = NA

colnames(gdp_popu)[1] <- "COUNTRY"
gdp_popu$COUNTRY <- toupper(gdp_popu$COUNTRY)
country_eq$Year <- format(country_eq$DATE,"%Y")
df_eq_gdp <- left_join(country_eq,gdp_popu)

colnames(df_eq_gdp)[48] <- "GDP_constant2010USD"
colnames(df_eq_gdp)[49] <- "GDP_currentUSD"
colnames(df_eq_gdp)[50] <- "GDP_percap_2010USD"
colnames(df_eq_gdp)[51] <- "GDP_percap_currentUSD"

```
- No decrease in the total deaths/damages in large earthquakes

```{r, message = FALSE, warning = FALSE, echo=FALSE, fig.height = 5, fig.width = 10}

fig1 <- ggplot(country_eq, aes(x= format(DATE,"%Y"), y = log1p(TOTAL_DEATHS))) +
  stat_summary(fun.y=sum,geom="bar") + 
  theme(axis.text.x = element_text(angle = 90, size = 8), axis.title =   element_text(size = 10)) +
  scale_x_discrete("Year") + 
  scale_y_continuous("log1p(tot. deaths)")

fig2 <- ggplot(country_eq, aes(x= format(DATE,"%Y"), y = log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS))) +
  stat_summary(fun.y=sum,geom="bar") + 
  theme(axis.text.x = element_text(angle = 90, size = 8), axis.title = element_text(size = 10)) +
  scale_x_discrete("Year") + 
  scale_y_continuous("log1p(tot. damages in M. dollars)") 

grid.arrange(fig1, fig2, ncol = 2)

```

## Visualization of the Data
- Positive correlation of earthquake magnitude and the total deaths/damages in an earthquake.

```{r, message = FALSE,warning = FALSE, echo = FALSE, fig.height = 2.5, fig.align = "center"}
f1 <- ggplot(country_eq,aes(x=EQ_PRIMARY, y = log1p(TOTAL_DEATHS))) +
  geom_point() +
  scale_x_continuous("earthquake magnitude") +
  scale_y_continuous("log1p(tot. deaths)") +
  theme(axis.title = element_text(size = 10))

f2 <- ggplot(country_eq,aes(x=EQ_PRIMARY, y = log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS))) +
  geom_jitter() +
  scale_x_continuous("earthquake magnitude") +
  scale_y_continuous("log1p(tot. damages)") +
  theme(axis.title = element_text(size = 10))

grid.arrange(f1, f2, ncol = 2)
```

- More deaths/damages are correlated with shallower earthquakes.

```{r, message = FALSE, warning = FALSE, echo = FALSE, fig.height = 2.5, fig.align = "center"}
p1 <- ggplot(country_eq,aes(x=FOCAL_DEPTH, y = log1p(TOTAL_DEATHS),color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_color_gradientn(colors=rainbow(4)) +
  scale_y_continuous("log1p(tot. deaths)") +
  theme(legend.position = "none")
#  scale_x_continuous(limits = c(0, 300))

p2 <- ggplot(country_eq,aes(x=FOCAL_DEPTH, y = log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS),
                      color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_color_gradientn(colors=rainbow(4)) + 
  scale_y_continuous("log1p(tot. damages)") 

grid.arrange(p1, p2, ncol = 2, widths = 2:3)
```

## Visualization of the Data
- No clear correlation between the total loss and a country's population.
- It appears that countries with higher GDP values suffered fewer deaths and less damages.

```{r, message = FALSE, warning = FALSE, echo = FALSE, fig.height = 5}

# Total Deaths VS GDP 
a1 <- ggplot(df_eq_gdp,aes(x = as.numeric(GDP_currentUSD),y=log1p(TOTAL_DEATHS), color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_x_continuous("GDP current $US") +
  scale_y_continuous("log1p(tot. death)") +
  scale_color_gradientn(colors=rainbow(4)) + 
  theme(axis.title = element_text(size = 10))

# Total damages VS GDP 
a2 <- ggplot(df_eq_gdp,aes(x = as.numeric(GDP_currentUSD),
                           y=log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS), color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_x_continuous("GDP current $US") +
  scale_y_continuous("log1p(tot. damages)") +
  scale_color_gradientn(colors=rainbow(4)) + 
  theme(axis.title = element_text(size = 10))

# Total Deaths VS Population
a3 <- ggplot(df_eq_gdp,aes(x = as.numeric(Population),y=log1p(TOTAL_DEATHS), color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_x_continuous("Country Population") +
  scale_y_continuous("log1p(tot. death)") +
  scale_color_gradientn(colors=rainbow(4)) + 
  theme(axis.title = element_text(size = 10))

# Total damages VS Population
a4 <- ggplot(df_eq_gdp,aes(x = as.numeric(Population),
                           y=log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS), color = EQ_PRIMARY)) +
  geom_jitter() +
  scale_x_continuous("Country Population") +
  scale_y_continuous("log1p(tot. damages)") +
  scale_color_gradientn(colors=rainbow(4)) + 
  theme(axis.title = element_text(size = 10))

#extract legend and add to the bottom
grid_arrange_shared_legend <- function(...) {
    plots <- list(...)
    g <- ggplotGrob(plots[[1]] + theme(legend.position="bottom"))$grobs
    legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
    lheight <- sum(legend$height)
    grid.arrange(
        do.call(arrangeGrob, lapply(plots, function(x)
            x + theme(legend.position="none"))),
        legend,
        ncol = 1,
        heights = unit.c(unit(1, "npc") - lheight, lheight))
}

grid_arrange_shared_legend(a1, a3, a2, a4)

```


## Machine Learning Methods - Linear Regression Analysis 
- High residual and low R-squared value in the summary of this model suggest a bad fit. 

```{r, message = FALSE,warning = FALSE, echo = FALSE, size = "tiny"}
# Regression Analysis
df_eq_gdp$GDP_constant2010USD <- as.numeric(df_eq_gdp$GDP_constant2010USD)
df_eq_gdp$GDP_currentUSD <- as.numeric(df_eq_gdp$GDP_currentUSD)
df_eq_gdp$Population <- as.numeric(df_eq_gdp$Population)

model1 <- lm(TOTAL_DEATHS ~ FOCAL_DEPTH + EQ_PRIMARY + INTENSITY + LATITUDE + LONGITUDE,
           data = df_eq_gdp)
summary(model1)
```

## Machine Learning Methods - Linear Regression Analysis 
- Including country's GDP and population does not provide a good fit model.

```{r, message = FALSE,warning = FALSE, echo = FALSE}
model2 <- lm(TOTAL_DEATHS ~ FOCAL_DEPTH + EQ_PRIMARY + INTENSITY + LATITUDE + LONGITUDE
             + GDP_currentUSD + Population,data = df_eq_gdp)
summary(model2)
```

## Machine Learning Methods - Random Forest 
- Large mean of squared residuals and the negative value for the % of variance indicates that random forest method was not adequate to define a good fit model.  

```{r, message = FALSE, warning = FALSE, echo = FALSE}
set.seed(130)
fit <- randomForest(TOTAL_DEATHS ~ FOCAL_DEPTH + EQ_PRIMARY + 
                      INTENSITY + LATITUDE + 
                      LONGITUDE + GDP_currentUSD + Population,
                    data = df_eq_gdp,
                    importance = TRUE,
                    na.action = na.omit,
                    ntree = 2000)
#varImpPlot(fit,type=1)
importance(fit)

fit
```

## Map the Earthquakes 
- Total deaths and total damages was plotted every 10 years, and at certain regions (the Mediterranean and South America), the total loss has decreased over the years.  

```{r, message = FALSE,warning = FALSE, echo = FALSE}
map.world <- map_data(map = "world")
my_map <- ggplot(map.world, aes(x = long, y = lat, group = group))
my_map <- my_map + geom_path(size = 0.2)

data_67_77 <- filter(df_eq_gdp, Year < 1977 & Year >= 1967)
data_77_87 <- filter(df_eq_gdp, Year < 1987 & Year >= 1977)
data_87_97 <- filter(df_eq_gdp, Year < 1997 & Year >= 1987)
data_97_07 <- filter(df_eq_gdp, Year < 2007 & Year >= 1997)
data_07_17 <- filter(df_eq_gdp, Year <= 2017 & Year >= 2007)

p1 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DEATHS)),
                      data = data_67_77, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
p1 <- p1 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
           scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
p1 <- p1 + labs(title = "1967 - 1977")

p2 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DEATHS)),
                          data = data_77_87, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
p2 <- p2 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
p2 <- p2 + labs(title = "1977 - 1987")

p3 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DEATHS)),
                          data = data_87_97, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
p3 <- p3 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
p3 <- p3 + labs(title = "1987 - 1997")

p4 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DEATHS)),
                          data = data_97_07, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
p4 <- p4 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
p4 <- p4 + labs(title = "1997 - 2007")

p5 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DEATHS)),
                          data = data_07_17, shape=21, stroke=0.5, alpha = 0.6, color="black", inherit.aes = FALSE)
p5 <- p5 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + 
  theme(legend.title = element_text(size = 8),legend.box = "horizontal", 
        legend.position = c(1.3, 0.5), plot.title = element_text(hjust = 0.5))
p5 <- p5 + labs(title = "2007 - 2017", fill = "log(tot. deaths)" )


#fig <- arrangeGrob(p1, p2, p3, p4, p5, ncol = 3)
#ggsave(file = "eq_death_plot.pdf", fig, width = 10, height = 6)
grid.arrange(p1, p2, p3, p4, p5, ncol = 3, heights=unit(c(2.5,2.5), c("in", "in")), newpage=TRUE)

```

## Map the Earthquakes 

```{r, message = FALSE,warning = FALSE, echo = FALSE}
# Plot the total damages every 10 years
d1 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS)),
                          data = data_67_77, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
d1 <- d1 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
d1 <- d1 + labs(title = "1967 - 1977")

d2 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS)),
                          data = data_77_87, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
d2 <- d2 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
d2 <- d2 + labs(title = "1977 - 1987")

d3 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS)),
                          data = data_87_97, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
d3 <- d3 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
d3 <- d3 + labs(title = "1987 - 1997")

d4 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS)),
                          data = data_97_07, shape=21, stroke=0.5, alpha = 0.6,color="black", inherit.aes = FALSE)
d4 <- d4 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + theme(legend.position="none", plot.title = element_text(hjust = 0.5))
d4 <- d4 + labs(title = "1997 - 2007")

d5 <- my_map + geom_point(aes(x = LONGITUDE, y = LATITUDE, size = EQ_PRIMARY,  fill= log1p(TOTAL_DAMAGE_MILLIONS_DOLLARS)),
                          data = data_07_17, shape=21, stroke=0.5, alpha = 0.6, color="black", inherit.aes = FALSE)
d5 <- d5 + scale_fill_continuous(low='yellow', high='red',na.value = NA) + 
  scale_size(range = c(0.2,3)) + 
  theme(legend.title = element_text(size = 8),legend.box = "horizontal", 
        legend.position = c(1.5, 0.5), plot.title = element_text(hjust = 0.5))
d5 <- d5 + labs(title = "2007 - 2017", fill = "log(tot. damages)")

grid.arrange(d1, d2, d3, d4, d5, ncol = 3, heights=unit(c(2.5,2.5), c("in", "in")), newpage=TRUE)
#fig2 <- arrangeGrob(d1, d2, d3, d4, d5, ncol = 3)
#ggsave(file = "eq_damage_plot.pdf", fig2, width = 10, height = 6)
```

## Results
- The overall total loss in large earthquakes did not decrease over the years, but the loss seemed to decrease at certain regions.
- Large earthquakes and shallow earthquakes are correlated with more damages and deaths. 
- These correlation, however, are not linear, as evidenced by the poor results from the linear regression analysis. 
- Preliminary inspection seems to show that rich countries (higher GDP) have less loss in earthquakes.
- The correlation between GDP and loss in earthquakes requires further investigation, becase earthquakes only struck certain areas while national GDP evaluates the economic status of an entire country. 

## Implications
- Results will be useful for the UN to budget emergency fund for earthquake-prone areas. 
- E.g., Areas that suffer great loss during the past 50 years and also have high risk for earthquakes in the future, a greater amount of emergency fund should be budgeted. 
- The analyses can be furthered to study similar correlations in individual country, state or even city. 
- Using these results to identify countries, states and cities that have successfully reduced the total loss over the years and apply their strategies to other regions. 